Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-08 conclusions.tex
Adding GW macro
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@974 ccdd05d5-56c1-461a-bf98-85f6876523b4

We have demonstrated a computationally feasible procedure for the rapid and
even preemptive detection of gravitational waves emitted during the coalescence
of neutron stars and stellar-mass black holes. Our method is as fast as
standard \fft{} convolutions but allows for zero latency, sample-in-sample-out
operation.  Our search targets are expected to produce prompt electromagnetic
signals and may be the progenitors of some short hard gamma-ray bursts.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  We anticipate requiring $\sim100$ modern multi-core computers to
analyze a four-detector network of gravitational-wave data for binary neutron
stars and stellar-mass black holes.  This is well within the current computing
capabilities of the \LIGO{} Data Grid~\cite{LDG}. In the future, this could be
improved upon further by conditionally reconstructing the \SNR{} time-series
only during times when a composite dection statistic crosses a
threshold~\cite{svd-compdetstat}.

>>>>

We have demonstrated a computationally feasible procedure for the rapid and
even preemptive detection of \GW{}s emitted during the coalescence
of neutron stars and stellar-mass black holes. Our method is as fast as
standard \fft{} convolutions but allows for zero latency, sample-in-sample-out
operation.  Our search targets are expected to produce prompt electromagnetic
signals and may be the progenitors of some short hard gamma-ray bursts.  Rapid
alerts to the broader astronomical community will improve the chances of
detecting an electromagnetic counterpart in bands from gamma-rays down to
radio.  We anticipate requiring $\sim100$ modern multi-core computers to
analyze a four-detector network of \GW{} data for binary neutron
stars and stellar-mass black holes.  This is well within the current computing
capabilities of the \LIGO{} Data Grid~\cite{LDG}. In the future, this could be
improved upon further by conditionally reconstructing the \SNR{} time-series
only during times when a composite dection statistic crosses a
threshold~\cite{svd-compdetstat}.

### 2011-06-08 implementation.tex
Adding GW macro
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@974 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe an implementation of the \lloid\ method described
in section \ref{sec:method} suitable for rapid gravitational-wave
searches for \CBC{}s.  The \lloid\ method requires several
computations that can be completed before the analysis is underway.  Thus
we divide the procedure into two stages 1) an offline planning stage and 2) an
online, low-latency filtering stage.  The offline stage can be done before the
analysis is started and updated asynchronously, whereas the online stage must
keep up with the detector output and produce search results as rapidly as
possible.  In the next two subsections we describe what these stages entail.

@@

The \lloid\ algorithm could be used in a true sample-in-sample-out real-time
system.  However, such a system would likely require integration directly into
the data acquisition and storage system of the gravitational-wave
observatories.  A slightly more modest goal is to leverage existing low-%
latency, but not real-time, signal processing infrastructure in order to
implement the \lloid\ algorithm.  For the near-term this should be a viable 
solution for searches with order seconds of intrinsic latency.

We have implemented a prototype of the low-latency filtering stage using an
open-source signal processing environment called \gstreamer\ \cite{gstreamer}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of
gravitational-wave detector data to audio data it is not surprising that
\gstreamer\ is useful for our purpose. \gstreamer\ also provides some useful
stock signal processing elements such as resamplers and filters.  We have
extended the \gstreamer\ framework by developing a library called
\gstlal{}~\cite{gstlal} that provides elements for gravitational-wave data
analysis.

>>>>

In this section we describe an implementation of the \lloid\ method described
in section \ref{sec:method} suitable for rapid \GW{}
searches for \CBC{}s.  The \lloid\ method requires several
computations that can be completed before the analysis is underway.  Thus
we divide the procedure into two stages 1) an offline planning stage and 2) an
online, low-latency filtering stage.  The offline stage can be done before the
analysis is started and updated asynchronously, whereas the online stage must
keep up with the detector output and produce search results as rapidly as
possible.  In the next two subsections we describe what these stages entail.

@@

The \lloid\ algorithm could be used in a true sample-in-sample-out real-time
system.  However, such a system would likely require integration directly into
the data acquisition and storage system of the \GW{}
observatories.  A slightly more modest goal is to leverage existing low-%
latency, but not real-time, signal processing infrastructure in order to
implement the \lloid\ algorithm.  For the near-term this should be a viable 
solution for searches with order seconds of intrinsic latency.

We have implemented a prototype of the low-latency filtering stage using an
open-source signal processing environment called \gstreamer\ \cite{gstreamer}.
\gstreamer\ is a vital component of many Linux systems, providing media
playback, authoring, and streaming on devices from cell phones to desktop
computers to streaming media servers.  Given the similarities of
\GW{} detector data to audio data it is not surprising that
\gstreamer\ is useful for our purpose. \gstreamer\ also provides some useful
stock signal processing elements such as resamplers and filters.  We have
extended the \gstreamer\ framework by developing a library called
\gstlal{}~\cite{gstlal} that provides elements for \GW{} data
analysis.

### 2011-06-08 introduction.tex
Adding GW macro
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@974 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves, its orbital
separation decays, leading to a run-away inspiral with the gravitational-wave
amplitude and frequency increasing until the system eventually merges soon
after the innermost stable circular orbit (\ISCO). If a neutron star is
involved it may become tidally disrupted near the merger.  This disrupted
matter can fuel a bright electromagnetic counterpart in the merger's
aftermath~\cite{shibata:2007}.

Gravitational-wave detectors are sensitive the to these gravitational-wave
signals.  Indeed, the coalescence of compact binary systems consisting of
neutron stars (\textsc{ns}) and/or black holes (\textsc{bh}) is the most
promising source of gravitational radiation for Advanced \LIGO, Virgo, \GEO\,
and \LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary
coalescence events per year are expected to be observed in the advanced
detector era later this decade~\cite{Abadie:2010p10836}. In the era of
first-generation detectors the gravitational-wave community initiated a project
to send alerts when potential gravitational-wave transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was an
important achievement, but far too late to catch prompt electromagnetic
emission.

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical
flash~\cite{Sari99}. Prompt emission is a probe into the extreme initial
conditions of the outflow, in contrast with afterglows, which arise in the
external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have only been observed to
peak within tens of seconds for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows. Rapid
gravitational-wave transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}. An electromagnetic counterpart would
vastly boost confidence in the gravitational-wave detection and provide the
tight sky localization necessary to allow determination of the source's host
galaxy, which leads to a redshift measurement. A coincident gravitational-wave
observation will provide a luminosity distance measurement. With these
together, we can produce precision measurements of the Hubble
constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting gravitational-wave candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By examining the signal to noise ratio (\SNR{}) stream
for threshold crossings before the gravitational-wave signal leaves the
detection band, it is possible to trade some \SNR{} and sky localization
accuracy~\cite{Fairhurst2009} for latency.  Figure \ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns} sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best gravitational-wave trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current gravitational-wave search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in gravitational-wave
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges soon
after the innermost stable circular orbit (\ISCO). If a neutron star is
involved it may become tidally disrupted near the merger.  This disrupted
matter can fuel a bright electromagnetic counterpart in the merger's
aftermath~\cite{shibata:2007}.

\GW{} detectors are sensitive the to these \GW{}
signals.  Indeed, the coalescence of compact binary systems consisting of
neutron stars (\textsc{ns}) and/or black holes (\textsc{bh}) is the most
promising source of gravitational radiation for Advanced \LIGO, Virgo, \GEO\,
and \LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary
coalescence events per year are expected to be observed in the advanced
detector era later this decade~\cite{Abadie:2010p10836}. In the era of
first-generation detectors the \GW{} community initiated a project
to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was an
important achievement, but far too late to catch prompt electromagnetic
emission.

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical
flash~\cite{Sari99}. Prompt emission is a probe into the extreme initial
conditions of the outflow, in contrast with afterglows, which arise in the
external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have only been observed to
peak within tens of seconds for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}. An electromagnetic counterpart would
vastly boost confidence in the \GW{} detection and provide the
tight sky localization necessary to allow determination of the source's host
galaxy, which leads to a redshift measurement. A coincident \GW{}
observation will provide a luminosity distance measurement. With these
together, we can produce precision measurements of the Hubble
constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting \GW{} candidates
not minutes \emph{after} the merger, as has already been accomplished, but
seconds \emph{before}.  By examining the signal to noise ratio (\SNR{}) stream
for threshold crossings before the \GW{} signal leaves the
detection band, it is possible to trade some \SNR{} and sky localization
accuracy~\cite{Fairhurst2009} for latency.  Figure \ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns} sources that will
be detectable $t$ seconds before coalescence.  The heavy solid line is the most
likely yearly rate estimate $\dot N_{\mathrm{re}}$ during Advanced \LIGO.  The
light solid lines represent the confidence interval $[\dot N_{\mathrm{low}}, \dot 
N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in 
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for
detection and that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a
detector in the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be 
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be detectable 
within 100 seconds of merger.  The `zero detuning, lower power' and `\textsc{bbh} optimized' configurations are still more conducive to advance detection
but at the cost of fewer total detectable events above an \SNR\ of 8.}
\end{center}
\end{figure}

In October 2010, \LIGO{} completed its sixth science run
(S6) and Virgo completed its third science run (VSR3).  While both
\LIGO{} detectors and Virgo were operating, several all-sky detection
pipelines operated in a low-latency configuration to send astronomical alerts,
namely \textsc{mbta}, Coherent WaveBurst, and
Omega~\cite{HugheyGWPAW2011, S6lowlatency}.
\editorial{Get references for these low-latency pipelines.} The S6 analyses
achieved latencies of 30--60 minutes, which were dominated by a human vetting
process. Candidates were sent for electromagnetic followup to several
telescopes; Swift, \textsc{rotse}, \textsc{tarot}, and Zadko~\cite{kanner2008,
HugheyGWPAW2011} took images of likely sky locations.  \textsc{mbta} achieved
the best \GW{} trigger-generation latencies of 2--5 minutes.  We
assume that in the advanced detector era the vetting process will be automated,
so current \GW{} search methodology and telescope actuation would
dominate latency.

Advance detection of compact binary coalescences (\CBC{}s) will require striking a balance between latency
and throughput.  \CBC{} searches consist of banks of matched filters, or
cross-correlations between the data stream and a bank of nominal ``template''
signals.  There are many different implementations of matched filters, but most
have high throughput at the cost of high latency, or low latency at the cost of
low throughput.  The former are epitomized by the overlap-save algorithm
\cite{numerical-recipes-chapter-13} for frequency domain (\FD) convolution,
currently the preferred method in \GW{}
searches.  The most obvious example of the latter is the time domain
(\TD) convolution, which has no algorithmic latency.  However, its
computational complexity is quadratic in the length of the templates, so it is
prohibitively expensive for long templates.

### 2011-06-08 method.tex
Adding GW macro
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@974 ccdd05d5-56c1-461a-bf98-85f6876523b4

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\vec\theta$ that determine the amplitude and phase evolution of the
gravitational wave strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\vec\theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\cite{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' gravitational wave polarizations.  Thus, for any value of
$\vec\theta$ we must implement 2 filters.  The coefficients for the $\numtmps$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$\numtmps/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\,
\vec\theta_{\numtmps/2-1}$ to assure a bounded loss of \SNR~\cite{Owen:1998dk}. That
is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $\numtmps$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.

>>>>

Inspiral signals are continuously parameterized by a set of intrinsic source
parameters $\vec\theta$ that determine the amplitude and phase evolution of the
\GW{} strain. For systems where the effects of spin can be
ignored, the intrinsic source parameters are the component masses of the
binary, $\vec\theta = (m_1, m_2)$. Searches for inspiral signals typically
employ matched filter banks that discretely sample the possible intrinsic
parameters~\cite{findchirppaper}.  For a given source, the strain observed by
the detector is a linear combination of two waveforms corresponding to the
`$+$' and `$\times$' \GW{} polarizations.  Thus, for any value of
$\vec\theta$ we must implement 2 filters.  The coefficients for the $\numtmps$ filters
are known as templates, and are formed by discretizing and time reversing the
waveforms and weighting them by the inverse amplitude spectral density of the
detector's noise. To construct a template bank, templates are chosen with the
$\numtmps/2$ discrete signal parameters $\vec\theta_0,\, \vec\theta_1,\, \dots,\,
\vec\theta_{\numtmps/2-1}$ to assure a bounded loss of \SNR~\cite{Owen:1998dk}. That
is, any possible signal within a given range of intrinsic
source parameters will have an inner product that is $\geqslant 0.97$ with at
least one template. Such a template bank is said to have a {\em minimum match}
of 0.97. The data from the detector are whitened and convolved with each
template to produce $\numtmps$ \SNR\ time series. Local peak-finding across time and
templates determines detection candidates.
